---
title: "AI Data Centers and Power Grid Load Shifts"
author: "Mingie Wei, Si Min Loo, Seohyun Oh"
date: "`r Sys.Date()`"
output: html_document
---

# Loading
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(tidyverse)
library(readxl)
library(lubridate)
library(dplyr)
```

```{r load-data}
tex_raw  <- read_excel("Region_TEX.xlsx")
pjm_raw <- read_excel("PJM.xlsx")

tex_weather <- read_csv("TEX_Weather.csv")
pjm_weather <- read_csv("PJM_Weather.csv")
```

# Cleaning them, keeping only demand, demand forecasted, and generation
```{r}

tex_clean <- tex_raw %>%
  select(`Local date`, Hour, `Demand forecast`, Demand, `Demand forecast`, `Net generation`) %>%
  rename(
    date            = `Local date`,
    hour            = Hour,
    demand_mwh      = Demand,
    demand_forecast = `Demand forecast`,
    generation = `Net generation`
  ) %>%
  mutate(date = as_date(date)) %>%
  filter(!is.na(demand_mwh))

pjm_clean <- pjm_raw %>%
  # Used Adjusted demand, Adjusted net generation for PJM 
  select(`Local date`, Hour, `Demand forecast`, `Adjusted demand`, `Demand forecast`, `Adjusted net generation`) %>%
  rename(
    date            = `Local date`,
    hour            = Hour,
    demand_mwh      = `Adjusted demand`,
    demand_forecast = `Demand forecast`,
    generation = `Adjusted net generation`
  ) %>%
  mutate(date = as_date(date)) %>%
  filter(!is.na(demand_mwh))
```

```{r}
tex_clean
pjm_clean
```


# aggregate to daily data
```{r}
tex_daily <- tex_clean %>%
  group_by(date) %>%
  summarise(
    demand_daily   = sum(demand_mwh, na.rm = TRUE),
    forecast_daily = sum(demand_forecast, na.rm = TRUE),
    gen_daily      = sum(generation, na.rm = TRUE),
    .groups = "drop"
  )

pjm_daily <- pjm_clean %>%
  group_by(date) %>%
  summarise(
    demand_daily   = sum(demand_mwh, na.rm = TRUE),
    forecast_daily = sum(demand_forecast, na.rm = TRUE),
    gen_daily      = sum(generation, na.rm = TRUE),
    .groups = "drop"
  )

range(pjm_daily$date)
range(tex_daily$date)

# Are there any gaps in the daily series?
tex_daily %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  filter(gap > 1)
pjm_daily %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  filter(gap > 1)
```

# Initial Exploratory plots
```{r}
bind_rows(
  pjm_daily %>% mutate(region = "PJM"),
  tex_daily  %>% mutate(region = "TEX")
) %>%
  ggplot(aes(x = date, y = demand_daily, color = region)) +
  geom_line(alpha = 0.3) +
  geom_smooth(method = "loess", span = 0.1, se = FALSE) +
  geom_vline(xintercept = as.Date("2022-11-01"),
             linetype = "dashed", color = "black") +
  annotate("text", x = as.Date("2022-11-01"),
           y = Inf, label = "ChatGPT launch", 
           hjust = -0.1, vjust = 1.5, size = 3) +
  facet_wrap(~region, scales = "free_y", ncol = 1) +
  labs(title = "Daily Electricity Demand",
       x = NULL, y = "Daily Demand (MWh)") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
bind_rows(
  pjm_daily %>% mutate(region = "PJM"),
  tex_daily  %>% mutate(region = "TEX")
) %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(region, month) %>%
  summarise(demand_monthly = mean(demand_daily), .groups = "drop") %>%
  ggplot(aes(x = month, y = demand_monthly, color = region)) +
  geom_line() +
  geom_vline(xintercept = as.Date("2022-11-01"),
             linetype = "dashed", color = "black") +
  facet_wrap(~region, scales = "free_y", ncol = 1) +
  labs(title = "Monthly Average Daily Demand",
       x = NULL, y = "Avg Daily Demand (MWh)") +
  theme_minimal() +
  theme(legend.position = "none")

bind_rows(
  pjm_daily %>% mutate(region = "PJM"),
  tex_daily  %>% mutate(region = "TEX")
) %>%
  mutate(year = year(date), doy = yday(date)) %>%
  filter(year >= 2018) %>%
  mutate(recent = ifelse(year >= 2022, "2022+", "Pre-2022")) %>%
  ggplot(aes(x = doy, y = demand_daily, 
             color = factor(year),
             alpha = recent,
             group = factor(year))) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE) +
  scale_alpha_manual(values = c("2022+" = 1, "Pre-2022" = 0.3)) +
  facet_wrap(~region, scales = "free_y", ncol = 1) +
  labs(title = "Demand by Day of Year, Colored by Year",
       subtitle = "Post-2022 years should sit above earlier years if break exists",
       x = "Day of Year", y = "Daily Demand (MWh)",
       color = "Year", alpha = NULL) +
  theme_minimal()


```

# PACF/ACF plots and STL Decomposition

```{r}
library(forecast)

# Check autocorrelation structure
par(mfrow = c(2,2))
acf(pjm_daily$demand_daily, lag.max = 365, main = "ACF - PJM")
pacf(pjm_daily$demand_daily, lag.max = 365, main = "PACF - PJM")
acf(tex_daily$demand_daily, lag.max = 365, main = "ACF - TEX")
pacf(tex_daily$demand_daily, lag.max = 365, main = "PACF - TEX")


pjm_ts <- ts(pjm_daily$demand_daily, 
             start = c(2016, 1), 
             frequency = 365)

tex_ts <- ts(tex_daily$demand_daily,
             start = c(2016, 1),
             frequency = 365)

pjm_stl <- stl(pjm_ts, s.window = "periodic", robust = TRUE)
tex_stl <- stl(tex_ts, s.window = "periodic", robust = TRUE)

plot(pjm_stl, main = "STL Decomposition - PJM")
plot(tex_stl, main = "STL Decomposition - TEX")

```
PJM: ACF shows a clear cyclic pattern with a 365-day lag, indicating strong annual seasonality.

TEX: The ACF displays a "U-shaped" decay, which is a classic symptom of a non-stationary series with a strong upward trend. This suggests that the mean demand is increasing over time, likely due to population growth and industrial expansion

# Merge Weather data from NOAA

```{r}
weather <- bind_rows(
  pjm_weather %>% mutate(region = "PJM"),
  tex_weather %>% mutate(region = "TEX")
) %>%
  mutate(
    DATE = as.Date(DATE),
    tavg = (TMAX + TMIN) / 2,
    cdd = pmax(tavg - 65, 0),
    hdd = pmax(65 - tavg, 0)
  )

weather_daily <- weather %>%
  group_by(DATE, region) %>%
  summarise(
    tavg = mean(tavg, na.rm = TRUE),
    cdd  = mean(cdd, na.rm = TRUE),
    hdd  = mean(hdd, na.rm = TRUE),
    .groups = "drop"
  )

pjm_final <- pjm_daily %>%
  mutate(date = as.Date(date)) %>%
  left_join(
    weather_daily %>%
      filter(region == "PJM") %>%
      select(-region),
    by = c("date" = "DATE")
  )

tex_final <- tex_daily %>%
  mutate(date = as.Date(date)) %>%
  left_join(
    weather_daily %>%
      filter(region == "TEX") %>%
      select(-region),
    by = c("date" = "DATE")
  )

pjm_final <- pjm_final %>%
  filter(!is.na(tavg), !is.na(cdd), !is.na(hdd))

tex_final <- tex_final %>%
  filter(!is.na(tavg), !is.na(cdd), !is.na(hdd))
```

# Train test split

# Baseline Modeling. Can try STL + Arima, Fourier, TBATS, some state space models?

# Do something about the structural break, maybe refit model after 2022 or something

# Metrics

# Compare across regions?







