---
title: "AI Data Centers and Power Grid Load Shifts"
author: "Mingie Wei, Si Min Loo, Seohyun Oh"
date: "`r Sys.Date()`"
output: html_document
---

# Loading
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(tidyverse)
library(readxl)
library(lubridate)
library(dplyr)
library(forecast)
```

```{r load-data}
tex_raw  <- read_excel("Region_TEX.xlsx")
pjm_raw <- read_excel("PJM.xlsx")
miso_raw <- read_excel("MISO.xlsx")

tex_weather <- read_csv("TEX_Weather.csv")
pjm_weather <- read_csv("PJM_Weather.csv")
miso_weather <- read_csv("MISO_Weather.csv")
```

# Cleaning them, keeping only demand, demand forecasted, and generation
```{r}

tex_clean <- tex_raw %>%
  select(`Local date`, Hour, `Demand forecast`, Demand, `Demand forecast`, `Net generation`) %>%
  rename(
    date            = `Local date`,
    hour            = Hour,
    demand_mwh      = Demand,
    demand_forecast = `Demand forecast`,
    generation = `Net generation`
  ) %>%
  mutate(date = as_date(date)) %>%
  filter(!is.na(demand_mwh))

pjm_clean <- pjm_raw %>%
  # Used Adjusted demand, Adjusted net generation for PJM 
  select(`Local date`, Hour, `Demand forecast`, `Adjusted demand`, `Demand forecast`, `Adjusted net generation`) %>%
  rename(
    date            = `Local date`,
    hour            = Hour,
    demand_mwh      = `Adjusted demand`,
    demand_forecast = `Demand forecast`,
    generation = `Adjusted net generation`
  ) %>%
  mutate(date = as_date(date)) %>%
  filter(!is.na(demand_mwh))

miso_clean <- miso_raw %>%
  # Used Adjusted demand, Adjusted net generation for PJM 
  select(`Local date`, Hour, `Demand forecast`, `Adjusted demand`, `Demand forecast`, `Adjusted net generation`) %>%
  rename(
    date            = `Local date`,
    hour            = Hour,
    demand_mwh      = `Adjusted demand`,
    demand_forecast = `Demand forecast`,
    generation = `Adjusted net generation`
  ) %>%
  mutate(date = as_date(date)) %>%
  filter(!is.na(demand_mwh))
```

```{r}
tex_clean
pjm_clean
miso_clean
```


# aggregate to daily data
```{r}
tex_daily <- tex_clean %>%
  group_by(date) %>%
  summarise(
    demand_daily   = sum(demand_mwh, na.rm = TRUE),
    forecast_daily = sum(demand_forecast, na.rm = TRUE),
    gen_daily      = sum(generation, na.rm = TRUE),
    .groups = "drop"
  )

pjm_daily <- pjm_clean %>%
  group_by(date) %>%
  summarise(
    demand_daily   = sum(demand_mwh, na.rm = TRUE),
    forecast_daily = sum(demand_forecast, na.rm = TRUE),
    gen_daily      = sum(generation, na.rm = TRUE),
    .groups = "drop"
  )

miso_daily <- miso_clean %>%
  group_by(date) %>%
  summarise(
    demand_daily   = sum(demand_mwh, na.rm = TRUE),
    forecast_daily = sum(demand_forecast, na.rm = TRUE),
    gen_daily      = sum(generation, na.rm = TRUE),
    .groups = "drop"
  )

range(pjm_daily$date)
range(tex_daily$date)
range(miso_daily$date)

# Are there any gaps in the daily series?
tex_daily %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  filter(gap > 1)
pjm_daily %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  filter(gap > 1)
miso_daily %>%
  mutate(gap = as.numeric(date - lag(date))) %>%
  filter(gap > 1)
```

# Initial Exploratory plots
```{r}
bind_rows(
  pjm_daily %>% mutate(region = "PJM"),
  tex_daily  %>% mutate(region = "TEX")
) %>%
  ggplot(aes(x = date, y = demand_daily, color = region)) +
  geom_line(alpha = 0.3) +
  geom_smooth(method = "loess", span = 0.1, se = FALSE) +
  geom_vline(xintercept = as.Date("2022-11-01"),
             linetype = "dashed", color = "black") +
  annotate("text", x = as.Date("2022-11-01"),
           y = Inf, label = "ChatGPT launch", 
           hjust = -0.1, vjust = 1.5, size = 3) +
  facet_wrap(~region, scales = "free_y", ncol = 1) +
  labs(title = "Daily Electricity Demand",
       x = NULL, y = "Daily Demand (MWh)") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
bind_rows(
  pjm_daily %>% mutate(region = "PJM"),
  tex_daily  %>% mutate(region = "TEX")
) %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(region, month) %>%
  summarise(demand_monthly = mean(demand_daily), .groups = "drop") %>%
  ggplot(aes(x = month, y = demand_monthly, color = region)) +
  geom_line() +
  geom_vline(xintercept = as.Date("2022-11-01"),
             linetype = "dashed", color = "black") +
  facet_wrap(~region, scales = "free_y", ncol = 1) +
  labs(title = "Monthly Average Daily Demand",
       x = NULL, y = "Avg Daily Demand (MWh)") +
  theme_minimal() +
  theme(legend.position = "none")

bind_rows(
  pjm_daily %>% mutate(region = "PJM"),
  tex_daily  %>% mutate(region = "TEX")
) %>%
  mutate(year = year(date), doy = yday(date)) %>%
  filter(year >= 2018) %>%
  mutate(recent = ifelse(year >= 2022, "2022+", "Pre-2022")) %>%
  ggplot(aes(x = doy, y = demand_daily, 
             color = factor(year),
             alpha = recent,
             group = factor(year))) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE) +
  scale_alpha_manual(values = c("2022+" = 1, "Pre-2022" = 0.3)) +
  facet_wrap(~region, scales = "free_y", ncol = 1) +
  labs(title = "Demand by Day of Year, Colored by Year",
       subtitle = "Post-2022 years should sit above earlier years if break exists",
       x = "Day of Year", y = "Daily Demand (MWh)",
       color = "Year", alpha = NULL) +
  theme_minimal()


```

# PACF/ACF plots and STL Decomposition

```{r}
# Check autocorrelation structure
par(mfrow = c(2,2))
acf(pjm_daily$demand_daily, lag.max = 365, main = "ACF - PJM")
pacf(pjm_daily$demand_daily, lag.max = 365, main = "PACF - PJM")
acf(tex_daily$demand_daily, lag.max = 365, main = "ACF - TEX")
pacf(tex_daily$demand_daily, lag.max = 365, main = "PACF - TEX")


pjm_ts <- ts(pjm_daily$demand_daily, 
             start = c(2016, 1), 
             frequency = 365)

tex_ts <- ts(tex_daily$demand_daily,
             start = c(2016, 1),
             frequency = 365)

pjm_stl <- stl(pjm_ts, s.window = "periodic", robust = TRUE)
tex_stl <- stl(tex_ts, s.window = "periodic", robust = TRUE)

plot(pjm_stl, main = "STL Decomposition - PJM")
plot(tex_stl, main = "STL Decomposition - TEX")

```
PJM: ACF shows a clear cyclic pattern with a 365-day lag, indicating strong annual seasonality.

TEX: The ACF displays a "U-shaped" decay, which is a classic symptom of a non-stationary series with a strong upward trend. This suggests that the mean demand is increasing over time, likely due to population growth and industrial expansion

# Merge Weather data from NOAA

```{r}
weather <- bind_rows(
  pjm_weather %>% mutate(region = "PJM"),
  tex_weather %>% mutate(region = "TEX"),
  miso_weather %>% mutate(region = "MISO")
) %>%
  mutate(
    DATE = as.Date(DATE),
    tavg = (TMAX + TMIN) / 2,
    cdd = pmax(tavg - 65, 0),
    hdd = pmax(65 - tavg, 0)
  )

weather_daily <- weather %>%
  group_by(DATE, region) %>%
  summarise(
    tavg = mean(tavg, na.rm = TRUE),
    cdd  = mean(cdd, na.rm = TRUE),
    hdd  = mean(hdd, na.rm = TRUE),
    .groups = "drop"
  )

pjm_final <- pjm_daily %>%
  mutate(date = as.Date(date)) %>%
  left_join(
    weather_daily %>%
      filter(region == "PJM") %>%
      select(-region),
    by = c("date" = "DATE")
  )

tex_final <- tex_daily %>%
  mutate(date = as.Date(date)) %>%
  left_join(
    weather_daily %>%
      filter(region == "TEX") %>%
      select(-region),
    by = c("date" = "DATE")
  )

miso_final <- miso_daily %>%
  mutate(date = as.Date(date)) %>%
  left_join(
    weather_daily %>%
      filter(region == "MISO") %>%
      select(-region),
    by = c("date" = "DATE")
  )

pjm_final <- pjm_final %>%
  filter(!is.na(tavg), !is.na(cdd), !is.na(hdd))

tex_final <- tex_final %>%
  filter(!is.na(tavg), !is.na(cdd), !is.na(hdd))

miso_final <- miso_final %>%
  filter(!is.na(tavg), !is.na(cdd), !is.na(hdd))
```

# Train test split

```{r}
# Hold out the last 365 days (~1 year) as a test set.

split_date <- max(tex_final$date) - 365

tex_train <- tex_final %>% filter(date <= split_date)
tex_test  <- tex_final %>% filter(date >  split_date)

pjm_train <- pjm_final %>% filter(date <= split_date)
pjm_test  <- pjm_final %>% filter(date >  split_date)

# Convert training series to ts (frequency = 365 approximates daily seasonality)
tex_train_ts <- ts(tex_train$demand_daily,
                   start     = c(year(min(tex_train$date)), yday(min(tex_train$date))),
                   frequency = 365)

pjm_train_ts <- ts(pjm_train$demand_daily,
                   start     = c(year(min(pjm_train$date)), yday(min(pjm_train$date))),
                   frequency = 365)

h <- nrow(tex_test)   # forecast horizon = length of test set

```

# Baseline Modeling. Can try STL + Arima, Fourier, TBATS, some state space models?

```{r}
# Reusable accuracy table builder
acc_table <- function(fc, actual, model_name, region) {
  e   <- actual - fc
  mae <- mean(abs(e))
  rmse <- sqrt(mean(e^2))
  mape <- mean(abs(e / actual)) * 100
  data.frame(region = region, model = model_name,
             MAE = round(mae, 0), RMSE = round(rmse, 0), MAPE = round(mape, 2))
}

#Model 1: SNaive
snaive_tex <- snaive(tex_train_ts, h = h)
snaive_pjm <- snaive(pjm_train_ts, h = h)
acc_snaive_tex <- acc_table(as.numeric(snaive_tex$mean), tex_test$demand_daily, "SNaive", "TEX")
acc_snaive_pjm <- acc_table(as.numeric(snaive_pjm$mean), pjm_test$demand_daily, "SNaive", "PJM")

# model 2: STL + ETS
stlf_tex <- stlf(tex_train_ts, h = h, method = "ets", robust = TRUE)
stlf_pjm <- stlf(pjm_train_ts, h = h, method = "ets", robust = TRUE)

acc_stlf_tex <- acc_table(as.numeric(stlf_tex$mean), tex_test$demand_daily, "STL+ETS", "TEX")
acc_stlf_pjm <- acc_table(as.numeric(stlf_pjm$mean), pjm_test$demand_daily, "STL+ETS", "PJM")

# STL + ARIMA
stlfarima_tex <- stlf(tex_train_ts, h = h, method = "arima", robust = TRUE)
stlfarima_pjm <- stlf(pjm_train_ts, h = h, method = "arima", robust = TRUE)
acc_stlfarima_tex <- acc_table(as.numeric(stlfarima_tex$mean), tex_test$demand_daily, "STL+ARIMA", "TEX")
acc_stlfarima_pjm <- acc_table(as.numeric(stlfarima_pjm$mean), pjm_test$demand_daily, "STL+ARIMA", "PJM")

# Fourier + ARIMA
K_tex <- 10   # initial, can do grid search later
K_pjm <- 10 # initial, can do grid search later

fourier_tex_train <- fourier(tex_train_ts, K = K_tex)
fourier_tex_test  <- fourier(tex_train_ts, K = K_tex, h = h)

fourier_pjm_train <- fourier(pjm_train_ts, K = K_pjm)
fourier_pjm_test  <- fourier(pjm_train_ts, K = K_pjm, h = h)

arima_fourier_tex <- auto.arima(tex_train_ts,
                                xreg        = fourier_tex_train,
                                seasonal    = FALSE,
                                stepwise    = TRUE,
                                approximation = TRUE)

arima_fourier_pjm <- auto.arima(pjm_train_ts,
                                xreg        = fourier_pjm_train,
                                seasonal    = FALSE,
                                stepwise    = TRUE,
                                approximation = TRUE)

fc_fourier_tex <- forecast(arima_fourier_tex, xreg = fourier_tex_test, h = h)
fc_fourier_pjm <- forecast(arima_fourier_pjm, xreg = fourier_pjm_test, h = h)

acc_fourier_tex <- acc_table(as.numeric(fc_fourier_tex$mean), tex_test$demand_daily, "Fourier+ARIMA", "TEX")
acc_fourier_pjm <- acc_table(as.numeric(fc_fourier_pjm$mean), pjm_test$demand_daily, "Fourier+ARIMA", "PJM")

# TBATS
tbats_tex <- tbats(tex_train_ts)
tbats_pjm <- tbats(pjm_train_ts)

fc_tbats_tex <- forecast(tbats_tex, h = h)
fc_tbats_pjm <- forecast(tbats_pjm, h = h)

acc_tbats_tex <- acc_table(as.numeric(fc_tbats_tex$mean), tex_test$demand_daily, "TBATS", "TEX")
acc_tbats_pjm <- acc_table(as.numeric(fc_tbats_pjm$mean), pjm_test$demand_daily, "TBATS", "PJM")

```

# Metrics

```{r}
# Reusable accuracy table builder

results <- bind_rows(
  acc_snaive_tex,    acc_snaive_pjm,
  acc_stlf_tex,      acc_stlf_pjm,
  acc_stlfarima_tex, acc_stlfarima_pjm,
  acc_fourier_tex,   acc_fourier_pjm,
  acc_tbats_tex,     acc_tbats_pjm
) %>%
  arrange(region, MAPE)

print(results)
```
# Plots 

```{r}
best_fc_tex <- stlfarima_tex   # ← update after comparing results

tibble(
  date     = tex_test$date,
  actual   = tex_test$demand_daily,
  forecast = as.numeric(best_fc_tex$mean),
  lo80     = as.numeric(best_fc_tex$lower[, 1]),
  hi80     = as.numeric(best_fc_tex$upper[, 1])
) %>%
  ggplot(aes(x = date)) +
  geom_ribbon(aes(ymin = lo80, ymax = hi80), fill = "steelblue", alpha = 0.2) +
  geom_line(aes(y = actual),   color = "black", linewidth = 0.7) +
  geom_line(aes(y = forecast), color = "steelblue", linewidth = 0.7) +
  labs(title  = "TEX – Best Model: Actual vs Forecast (Test Period)",
       subtitle = "Black = actual | Blue = forecast | Shaded = 80% CI",
       x = NULL, y = "Daily Demand (MWh)") +
  theme_minimal()

best_fc_pjm <- stlfarima_pjm   # ← update after comparing results

tibble(
  date     = pjm_test$date,
  actual   = pjm_test$demand_daily,
  forecast = as.numeric(best_fc_pjm$mean),
  lo80     = as.numeric(best_fc_pjm$lower[, 1]),
  hi80     = as.numeric(best_fc_pjm$upper[, 1])
) %>%
  ggplot(aes(x = date)) +
  geom_ribbon(aes(ymin = lo80, ymax = hi80), fill = "steelblue", alpha = 0.2) +
  geom_line(aes(y = actual),   color = "black", linewidth = 0.7) +
  geom_line(aes(y = forecast), color = "steelblue", linewidth = 0.7) +
  labs(title  = "PJM – Best Model: Actual vs Forecast (Test Period)",
       subtitle = "Black = actual | Blue = forecast | Shaded = 80% CI",
       x = NULL, y = "Daily Demand (MWh)") +
  theme_minimal()

```
# Deep Dive: Fourier + ARIMA with Weather Regressors

```{r}
run_demand_experiment <- function(full_data, train_cutoff, test_start, test_end, region_name, K_val = 5) {
  
  all_df <- full_data %>%
    arrange(date) %>%
    mutate(
      cdd_lag1 = lag(cdd, 1),
      hdd_lag1 = lag(hdd, 1),
      cdd_sq = cdd^2
    ) 
  
  train_df <- all_df %>%
  filter(date <= train_cutoff) %>%
  filter(complete.cases(cdd, hdd, cdd_lag1, hdd_lag1, cdd_sq, demand_daily))

  test_df <- all_df %>%
  filter(date >= test_start & date <= test_end) %>%
  filter(complete.cases(cdd, hdd, cdd_lag1, hdd_lag1, cdd_sq, demand_daily))
  
  weather_vars <- c("cdd", "hdd", "cdd_lag1", "hdd_lag1", "cdd_sq")
  
  train_x_raw <- as.matrix(train_df[, weather_vars])
  test_x_raw  <- as.matrix(test_df[, weather_vars])
  
  train_scaled <- scale(train_x_raw)
  test_scaled  <- scale(test_x_raw, 
                        center = attr(train_scaled, "scaled:center"), 
                        scale  = attr(train_scaled, "scaled:scale"))
  
  train_ts <- ts(train_df$demand_daily, frequency = 365.25)
  f_train  <- fourier(train_ts, K = K_val)
  f_test   <- fourier(train_ts, K = K_val, h = nrow(test_df))
  
  fit <- auto.arima(
    train_ts, 
    xreg = cbind(f_train, train_scaled), 
    seasonal = FALSE, stepwise = TRUE, approximation = FALSE
  )
  
  fc <- forecast(fit, xreg = cbind(f_test, test_scaled))
  
  res_table <- acc_table(
    as.numeric(fc$mean), 
    test_df$demand_daily, 
    paste0("Fourier+ARIMA+Weather (", year(train_cutoff-1),")"),
    region_name
  )
  
  return(list(table = res_table, model = fit, data = test_df %>% mutate(pred = as.numeric(fc$mean))))
}

tex_split_modern <- max(tex_final$date) - 365
pjm_split_modern <- max(pjm_final$date) - 365

tex_modern <- run_demand_experiment(
  full_data    = tex_final, 
  train_cutoff = tex_split_modern,        
  test_start   = tex_split_modern + 1,        
  test_end     = max(tex_final$date),        
  region_name  = "TEX"
)

pjm_modern <- run_demand_experiment(
  full_data    = pjm_final, 
  train_cutoff = pjm_split_modern, 
  test_start   = pjm_split_modern + 1, 
  test_end     = max(pjm_final$date), 
  region_name  = "PJM"
)

modern_results <- bind_rows(pjm_modern$table, tex_modern$table)
print(modern_results)

```

Both PJM and TEX significantly improved MAPE after controlling for weather

# Compare Modern (Trained pre-2025) vs Frozen (Trained pre-2022)

```{r}

cutoff_2022 <- as.Date("2022-11-01")

tex_modern <- run_demand_experiment(tex_final, tex_split_modern, tex_split_modern + 1, max(tex_final$date), "TEX")
pjm_modern <- run_demand_experiment(pjm_final, pjm_split_modern, pjm_split_modern + 1, max(pjm_final$date), "PJM")

tex_frozen <- run_demand_experiment(tex_final, cutoff_2022, tex_split_modern + 1, max(tex_final$date), "TEX")
pjm_frozen <- run_demand_experiment(pjm_final, cutoff_2022, pjm_split_modern + 1, max(pjm_final$date), "PJM")

final_comparison_table <- bind_rows(
  pjm_modern$table, tex_modern$table,
  pjm_frozen$table, tex_frozen$table
)

print(final_comparison_table)
```
The fact that the Frozen model (pre-2022) performs significantly worse, despite controlling for weather, proves that the historical relationship between weather and demand has been broken by a structural shift that didn't exist in 2022.

# Residual check to see the structural break

```{r}
checkresiduals(tex_modern$model)
checkresiduals(pjm_modern$model)
```
The Ljung-Box test (p-value < 2.2e-16) indicates that the residuals are not white noise, suggesting remaining autocorrelation not captured by the model. This persistent autocorrelation implies that important drivers of electricity demand beyond weather and seasonality—such as economic activity or infrastructure expansion—are not fully captured.

```{r}
comparison_data <- bind_rows(
  tex_modern$data %>% mutate(model_type = "Modern (Trained until 2025)", region = "TEX"),
  tex_frozen$data %>% mutate(model_type = "Frozen (Trained until 2022)", region = "TEX"),
  pjm_modern$data %>% mutate(model_type = "Modern (Trained until 2025)", region = "PJM"),
  pjm_frozen$data %>% mutate(model_type = "Frozen (Trained until 2022)", region = "PJM")
) %>%
  mutate(residual = demand_daily - pred)

ggplot(comparison_data, aes(x = residual, fill = model_type)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~region, scales = "free") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residual Density: Modern vs Frozen",
       subtitle = "A right-shift in Frozen model indicates persistent under-prediction due to structural change",
       x = "Residual (Actual - Predicted, MWh)", y = "Density") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplot(comparison_data, aes(x = date, y = residual, color = model_type)) +
  geom_smooth(method = "gam", se = TRUE) + 
  facet_wrap(~region, scales = "free_y") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residual Trend Over Time (2025-2026)",
       subtitle = "The gap between Actual and Frozen-model predictions represents the 'New Demand'",
       x = "Date", y = "Residual (MWh)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```
The massive, persistent right-shift in the Frozen model's residuals, especially in the TEX region—provides empirical evidence of a structural break. This confirms that a new baseline of electricity demand has emerged post-2022 that weather and seasonality alone can no longer explain.

# Compare across regions (MISO vs TEX)

```{r}
# Check autocorrelation structure
par(mfrow = c(1,2))
acf(miso_daily$demand_daily, lag.max = 365, main = "ACF - MISO")
pacf(miso_daily$demand_daily, lag.max = 365, main = "PACF - MISO")


miso_ts <- ts(miso_daily$demand_daily, 
             start = c(2016, 1), 
             frequency = 365)

miso_stl <- stl(miso_ts, s.window = "periodic", robust = TRUE)

plot(miso_stl, main = "STL Decomposition - MISO")
```
MISO: ACF shows a clear cyclic pattern with a 365-day lag, indicating strong annual seasonality.

```{r}
miso_split_modern <- max(miso_final$date) - 365

miso_modern <- run_demand_experiment(
  full_data    = miso_final, 
  train_cutoff = miso_split_modern,        
  test_start   = miso_split_modern + 1,        
  test_end     = max(miso_final$date),        
  region_name  = "MISO"
)

miso_frozen <- run_demand_experiment(miso_final, cutoff_2022, miso_split_modern + 1, max(miso_final$date), "MISO")

final_comparison_table_miso <- bind_rows(
  miso_modern$table, miso_frozen$table,
)

print(final_comparison_table_miso)
```

```{r}
checkresiduals(miso_modern$model)
```
```{r}
comparison_data2 <- bind_rows(
  tex_modern$data %>% mutate(model_type = "Modern (Trained until 2025)", region = "TEX"),
  tex_frozen$data %>% mutate(model_type = "Frozen (Trained until 2022)", region = "TEX"),
  miso_modern$data %>% mutate(model_type = "Modern (Trained until 2025)", region = "MISO"),
  miso_frozen$data %>% mutate(model_type = "Frozen (Trained until 2022)", region = "MISO")
) %>%
  mutate(residual = demand_daily - pred)

ggplot(comparison_data2, aes(x = residual, fill = model_type)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~region, scales = "free") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residual Density: Modern vs Frozen",
       subtitle = "A right-shift in Frozen model indicates persistent under-prediction due to structural change",
       x = "Residual (Actual - Predicted, MWh)", y = "Density") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplot(comparison_data2, aes(x = date, y = residual, color = model_type)) +
  geom_smooth(method = "gam", se = TRUE) + 
  facet_wrap(~region, scales = "free_y") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residual Trend Over Time (2025-2026)",
       subtitle = "The gap between Actual and Frozen-model predictions represents the 'New Demand'",
       x = "Date", y = "Residual (MWh)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```
There is no strong evidence of a persistent increase in demand in MISO; instead, the changes appear to be temporal and seasonal in nature. In contrast, TEX exhibits a systematic upward shift in residuals, indicating a persistent increase in baseline demand. 













